### 2020-01-24 Writing/disparity_review.tex
Updated latex

\textbf{The disparate ways to analyse morphological disparity}

Thomas
Guillerme\includegraphics[width=0.4902in,height=0.1807in]{./ObjectReplacements/Object 2},
Natalie
Cooper\includegraphics[width=0.3984in,height=0.1807in]{./ObjectReplacements/Object 4},
Stephen
Brusatte\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 6},
Katie
Davis\includegraphics[width=0.1992in,height=0.1807in]{./ObjectReplacements/Object 8},
Andrew L
Jackson\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 10},
Sylvain
Gerber\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 12},
Anjali
Goswami\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 14},
Kevin
Healy\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 16},
Melanie
Hopkins\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 18},
Graeme
Lloyd\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 20},
Joseph
O'Reilly\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 22},
Abi
Pate\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 24},
Emilie
Rayfield\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 26},
Erin
Saupe\includegraphics[width=0.239in,height=0.1807in]{./ObjectReplacements/Object 28},
Emma
Sherratt\includegraphics[width=0.2425in,height=0.1807in]{./ObjectReplacements/Object 30},
Graham
Slater\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 32},
Gavin H
Thomas\includegraphics[width=0.2465in,height=0.1807in]{./ObjectReplacements/Object 34}
and Philip
Donoghue\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 36}\\
\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 38}\emph{School
of Biological Sciences, University of Queensland, St. Lucia, Queensland,
Australia.}\\
\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 40}\emph{Department
of Life Sciences, Natural History Museum, London, Cromwell Road, London,
SW7 5BD, UK.}\\
\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 42}\emph{Bristol}

@@

*\textbf{Corresponding author: }\emph{guillert@tcd.ie}\\

@@

\textbf{-Please write in Google docs using }\emph{\textbf{Suggesting
mode}}\textbf{ only }(upper right corner of your screen)

@@

-I suggest you add your references using the \textbf{Paperpile} addon
(go in Add-ons menu \textgreater{} Get Add-ons\ldots{} \textgreater{}
Search for Paperpile \textgreater{} Add/Install; then go in Add-ons
menus \textgreater{} Paperpile \textgreater{} Manage citation). If you
don't use Paperpile, please don't add references as just (``Someone et
al. 2000'' but add also a link to the correct Google Scholar entry in
comments).

\textbf{-Add your affiliation in the author list above}. The
affiliations count in the word count so please 1) keep it short (max 1
per person!) and 2) try to use one that's already in there. Also add any
initials I've missed in your names.

- If in doubt, check formatting here:
\href{https://royalsociety.org/journals/authors/author-guidelines/}{\emph{https://royalsociety.org/journals/authors/author-guidelines/}}

@@

Morphological disparity analysis is used to characterize and investigate
discrete patterns of variation in the shape, form, function, and the
ecology of organisms. Since their conception in the 1980s, these methods
have been used to ask a variety of questions about how and why
morphological diversity evolves, and the ecological consequences of
these changes. There are many methodological approaches for studying
disparity, however, all of these come with caveats and pitfalls that
can, if ignored, lead to misinterpretations of disparity patterns. Here
we review morphological disparity methods and their caveats, and provide
best practice guidelines for their application and interpretation. We
discuss the data required for disparity studies, as well as
methodological challenges, including multidimensionality, ordination,
metric choice, hypothesis testing, and ancestral state estimations.
There is no ``one-size-fits-all'' approach to characterizing and
investigating the evolution of morphological disparity, and the
available tools should always be used in the context of a specific
biological question that will determine data and method selection at
every stage of the analysis.

\textbf{Keywords}: multidimensionality, palaeobiology, ecology\\

1. Introduction

Variation in the shape, form and function of organisms often exhibits a
pattern of discrete clusters rather than continuous gradients. These
patterns can be pictured as multidimensional spaces encompassing all the
possible forms an organism can take, and where each individual region of
this space is the result of a unique combination of many variables. One
commonly used approach to analyse these patterns and to understand the
processes that lead to them, is to analyse the morphological diversity
of a group, often referred to as a ``morphological disparity analysis''
\href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al. 2018)}.

The initial informal definition of morphological disparity was
``multidimensional morphological dissimilarity at a macroevolutionary
scale'' (Runnegar 1987,
\href{https://paperpile.com/c/sTGYvp/Uns3}{(Gould 2000)}). In the
seminal disparity papers of the late 1980s, disparity was used to
investigate body plan evolution after the Cambrian explosion
\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
1992)}\href{https://paperpile.com/c/sTGYvp/CidX+6tNm+oenu+TtGs}{(Gould
1991; Valentine 1986; M. Foote and Gould 1992; Morris 1989)};
\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
1992)};. Early disparity analyses focused on broad questions such as:
How does diversity evolve? Are some phenotypes more common than others?
and Can anatomical shapes evolve in all ``directions'' of the
multidimensional space
\href{https://paperpile.com/c/sTGYvp/yqPw+fTJ3}{(Mike Foote 1997,
1995)}? Since the 1990s, a great deal of literature from a range of
sub-fields in biology and palaeontology, and an equally large number of
methods, have used ``disparity'' as a broader umbrella term for studies
of multivariate variation. Although most influential reviews have
treated disparity as a pattern
\href{https://paperpile.com/c/sTGYvp/yqPw+nFf7+vTHS}{(Mike Foote 1997;
Wills 2001; Melanie J. Hopkins and Gerber 2017)}, there is currently no
consensus on whether to restrict use of the term disparity to its
original macroevolutionary-scale definition, or whether it can refer to
any study of multivariate variation. Here we use a broader definition
and suggest that morphological disparity is a \emph{pattern of variation
}in phenotype. Disparity analysis has been used to tackle a vast array
of questions. We identify four main classes of disparity analysis as
follows.

(1) \emph{Descriptive disparity} \emph{analyses. }The seminal studies of
disparity described the shapes of organisms and how they differed among
groups \href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Mike Foote 1995;
Wills, Briggs, and Fortey 1994)}. These consist of descriptive
multidimensionality analyses of the patterns in the diversity of
morphological traits, addressing questions such as: why are some
morphological traits combinations more common than others and what are
the properties of the resulting morphospace?
\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Mike Foote 1995;
Raup 1961; Gerber 2017)}.

(2) \emph{Disparity-through-time} \emph{analyses. }This approach
investigates how the shapes of organisms have changed through time,
focussing on the disparity of taxa in particular time bins or slices.
This approach has been widely used in palaeobiology to answer a range of
macroevolutionary questions
\href{https://paperpile.com/c/sTGYvp/PbSx+khc9+xxh5}{(Close et al. 2015;
Cisneros and Ruta 2010; Hughes, Gerber, and Wills 2013)}, such as: how
does disparity accumulate over the history of a clade
\href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper
2018; Wright 2017)}, and how does disparity change across mass
extinction events \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman
2010)}?

(3) \emph{Disparity and diversity. }Morphological disparity provides
another perspective on biodiversity; high morphological disparity
represents a high diversity of morphologies (i.e. shapes or body plans)
and is, presumably, associated with high levels of ecological and
functional diversity (see below). This makes disparity a more useful
measure of diversity than merely species richness. Indeed, most studies
that have investigated disparity and taxonomic diversity support an
effective decoupling
\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(Fortey,
Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and
Neige 2007)} - though this result breaks down at extremes of diversity;
extremely low diversity tends to yield low disparity and \emph{vice
versa}. This approach has been used to investigate whether some groups
are more successful than other in their exploration of new evolutionary
strategies @@@.

(4) \emph{Disparity and ecology. }The disparity of a group can be used
as a proxy for either the functional role it plays within an ecosystem
or its ecological niche. These analyses assume that groups with high
disparity are also likely to be functionally and ecologically diverse,
and that groups found in similar regions of shape space will have
similar functional and ecological roles
\href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010;
Pierce, Angielczyk, and Rayfield 2008; P. S. L. Anderson et al. 2011)}.
Note however, that the links between form and function are not always
clear cut; traits can be linked to multiple functions and multiple
functions can be linked to a single trait
\href{https://paperpile.com/c/sTGYvp/Ejzr}{\emph{(Wainwright et al.
2005)}}. This approach is used to investigate competitive replacements
\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
2008)} and changes in ecosystem function during and after mass
extinctions \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}.
It is particularly common in paleobiology where it is not possible to
observe directly the ecological or functional characteristics of extinct
species. \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.
2005)}

Of course, these four broad categories are not independent and many
studies use a combination of approaches
\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
2008)}. Furthermore, disparity analysis is continuously expanding into
areas. While this is exciting for the field, there are several issues
that may reduce the utility of disparity analyses if they are not
tackled. Here we review these issues and present best practice
guidelines to allow researchers to optimise their analyses. We first
discuss the appropriate data required for disparity analyses, then
review various challenging methodological aspects of disparity analyses.
Throughout it is important to remember that these tools should always be
used in the context of a specific scientific question, as this will
drive data and methodological choices at every stage of the process.

\includegraphics[width=6.5in,height=4.9445in]{Pictures/10000201000002EE0000023BE5B78D4C4B79F709.png}

\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main
types of disparity analysis. Descriptive disparity analyses focus on
describing the properties of the morphospace ; Disparity-through-time
analyses investigate the evolution of the morphospace through time
including the effect of extinction events; Disparity and diversity
analysis use disparity as a proxy for species diversity; Disparity and
ecology analyses use disparity as a proxy for the ecological or
functional role of a group. These categories are not independent and
many analyses will cover more than one.}

2. Data in Disparity Analyses

Regardless of the precise approach being used, disparity analyses are
based on morphological traits. It is important, therefore, to consider
which traits to use in the context of the specific question under
investigation. Traits for disparity analyses come from three main types
of data: (1) discrete ``cladistic'' characters, i.e coding the
absence-or-presence of features or a discrete characteristic of a trait
\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}); (2)
continuous measurements of features
\href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al.
2011)}; or (3) more sophisticated mathematical measurements from
geometric morphometric data (e.g. Procrustes superimpositions
\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, or
Fourier contours \href{https://paperpile.com/c/sTGYvp/ZEDR}{(Spriggs et
al. 2018)}). None of these are fundamentally better, however, some
questions and methodological pipelines are more appropriate with
specific types of data.

When investigating variation within bat wing shapes, for example, both
homologous landmarks from geometric morphometric analysis and continuous
measurements of bones may be appropriate to capture patterns of wing
variation. However, if the question focuses on the convergence between
bat and bird wings, the homology of the traits to measure is more
complex to define and the outline shape of the wings might more readily
answer the question from a mechanical perspective. Similarly, if the
focus is on wing function, i.e. whether the aerodynamic properties of
wings vary within bats or between bats and birds, the traits collected
should reflect these aerodynamic properties (e.g. wingspan, aspect
ratio, etc.). Whereas if the focus is on convergence between different
bats and birds, it would be preferable to use traits that have
facilitated flight in both groups (e.g. digit length, integumentary
system, etc.). Where there is any doubt about the appropriate traits to
choose, we advise using different kinds of data for the same feature to
determine whether it captures the same disparity.

This dicussion assumes that researchers are collecting their own data
for disparity analyses, but often this is not the case. It is common,
especially in disparity-through-time analyses, to recycle discrete
character data collected for use in phylogenetic inference (e.g. REFs
that have done this). This has the advantage of being quick, cheap and
efficient, but tends to artificially increase disparity between
phylogenetically distinct groups because phylogenetic characters are
collected specifically for the purpose of discriminating groups
\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Mike Foote 1995)}. This
needs to be considered when interpreting results, especially as when new
clades appear, the synapomorphies of that clade will naturally lead to
an apparent shift or increase in disparity. Similarly, data used for
phylogenetic reconstruction is likely to be highly skewed towards easily
available and taxonomically relevant traits, for example teeth in
mammals. If your disparity question is related to feeding ecology or
similar then this may be appropriate, but if the focus is on dispersal
ability then it would be sensible to select only postcranial traits. It
is often tempting to use all of a large dataset to increase the power of
your analyses, but if the data are not suitable for the question at hand
then the results may be meaningless (although ssome studies have suggest
this may not be a major issue;
\href{https://paperpile.com/c/sTGYvp/ZqML+Qrba}{(Zou and Zhang 2016;
Dávalos et al. 2014)}
\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}).

Finally, trait data suffers from the same shortcomings as most data --
data can be missing, non-overlapping, hierarchical, inapplicable,
ambiguous, polymorphic, correlated, or have insufficient sample size
\href{https://paperpile.com/c/sTGYvp/yO2t}{(Palci and Lee 2018)},
\href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau, Guillerme, and
Smith 2017)} {[}String of morpho data problems cite{]}. Biological
phenomena such as allometry and sexual dimorphism may also influence
trait data. More practically, data collection is constrained by the time
and money available, making collating a ``perfect'' dataset difficult.
We suggest that trait data are collected with the question in mind and
considering all of the caveats above, but where issues are unavoidable
at the very least these caveats must be considered during the analyses
and subsequent interpretation of the results.

\emph{Figure data {[}figure\_data.pdf{]}: major routes to obtain
disparity data. Data can be collected as discrete trait observations
(e.g. presence or absence data) or as continuous data. Continuous data
can be collected by different methods: linear measurements, landmark
coordinates or contours (curves). These measurement can then in turn be
mathematically transformed (logarithm, scaling, Procrustes
Superimposition, Elliptic Fourier, etc.). Regardless of the method, this
data collection results in a trait matrix where the observed traits
constitute columns and the studied elements (e.g. taxa) the rows.}

\includegraphics[width=6.5in,height=6.5in]{Pictures/10000201000007D0000007D0F8A72A6590C667D3.png}

3. Disparity Methods

Once suitable trait data have been collected, we need to consider the
disparity analysis. Below we address several key aspects including (i)
the difficulty of dealing with multidimensional data; (ii) caveats
associated with using ordination techniques to reduce the dimensionality
of the data; (iii) various metrics used to summarise the disparity of a
group; (iv) caveats related to methods that exist for hypothesis testing
within a disparity analysis framework; and (v) the specific case of
ancestral state estimations in disparity-through-time analyses. As noted
above, these tools should always be used in the context of a specific
question, as this will determine which methods are most appropriate.

\protect\hypertarget{anchor}{}{}3.1. Multidimensional mayhem

>>>>

\textbf{Disparities in the analysis of morphological disparity}

Thomas Guillerme\({}^{1,*, +}\), Natalie Cooper\({}^{2, +}\), Stephen L.
Brusatte\({}^{3}\), Katie E. Davis\({}^{4}\), Andrew L
Jackson\({}^{5}\), Sylvain Gerber\({}^{6}\), Anjali Goswami\({}^{2}\),
Kevin Healy\({}^{7}\), Melanie Hopkins\({}^{8}\), Marc EH
Jones\({}^{9}\), Graeme T. Lloyd\({}^{10}\), Joseph E.
O'Reilly\({}^{11}\), Abi Pate\({}^{11}\), Mark N Puttick\({}^{12}\),
Emiliy Rayfield\({}^{11}\), Erin E. Saupe\({}^{13}\), Emma
Sherratt\({}^{14}\), Graham Slater\({}^{15}\), Vera
Weisbecker\({}^{1}\), Gavin H Thomas\({}^{16}\) and Philip
Donoghue\({}^{11}\)\\
\({}^{1}\)\emph{School of Biological Sciences, University of Queensland,
St. Lucia, Queensland, Australia.}\\
\({}^{2}\)\emph{Department of Life Sciences, Natural History Museum,
London, Cromwell Road, London, SW7 5BD, UK.}

\emph{\textsuperscript{3}School of GeoSciences, University of Edinburgh,
Grant Institute, Edinburgh EH9 3FE, UK}

\textsuperscript{4} Department of Biology, University of York, YO10 5DD

\textsuperscript{5} Department of Zoology, School of Natural Sciences,
Trinity College.

\textsuperscript{6} Institut de Systématique, Évolution, Biodiversité
(ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne
Université, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005
Paris, France.

\textsuperscript{7} Ryan Institute, School of Natural Sciences, National
University of Ireland Galway, Ireland.

\textsuperscript{10} \emph{School of Earth and Environment, University
of Leeds, Leeds LS2 9JT, UK}\\
\textsuperscript{11}\emph{School of Earth Sciences, University of
Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK}

\textsuperscript{12} \emph{Milner Centre for Evolution, University of
Bath, BA2 7AY}

\textsuperscript{13} \emph{Department of Earth Sciences, University of
Oxford, S Parks Road, Oxford OX1 3AN, UK}

\textsuperscript{14} \emph{School of Biological Sciences, The University
of Adelaide, Adelaide, South Australia 5005, Australia}

@@

*\textbf{Corresponding author:} \emph{guillert@tcd.ie}

@@

@@

\textbf{-Add your affiliation in the author list above}. The
affiliations count in the word count so please 1) keep it short (max 1
per person!) and 2) try to use one that's already there. Also add any
initials I've missed in your names.

- If in doubt, check formatting here:
\href{https://royalsociety.org/journals/authors/author-guidelines/}{{https://royalsociety.org/journals/authors/author-guidelines/}}

@@

Statistical nalyses of morphological disparity have been used to
characterise and investigate the evolution of variation in the anatomy,
function, and ecology of organisms since the 1980s. While a diversity of
methods have been employed, it is unclear whether they provide
equivalent insights. Here we review the most commonly used approaches
for characterising and analysing morphological disparity, all of which
have associated limitations that, if ignored, can lead to
misinterpretation. We provide best practice guidelines for disparity
analyses, while noting that there can be no ``one-size-fits-all''
approach. The available tools should always be used in the context of a
specific biological question that will determine data and method
selection at every stage of the analysis.

\textbf{Keywords}: multidimensionality, palaeobiology, ecology,
morphology, disparity, variance/variation

\hypertarget{introduction}{%
\section{1. Introduction}\label{introduction}}

Clades of organisms are characterised by variation in both numbers of
species and range of phenotypes through time. At the extremes, clades
may be exceptionally species-rich and phenotypically diverse (e.g.
cichlids and molluscs), species-rich but phenotypically conservative
(e.g. bacteria and nematodes), species-poor but phenotypically diverse
(e.g. Afrotheria), or depauperate in both species and phenotypic
diversity (e.g. lungfish). These phenomena suggest that taxonomic and
phenotypic diversity are not inextricably linked, raising important
questions about how phenotypic diversity arises, such as: How does
diversity evolve? Are some morphologies more common than others? Can
anatomy evolve in all ``directions'' or are some anatomies impossible to
achieve? (Mike Foote 1997, 1995) What role does ecology play in
structuring morphological diversity?. Analyses of species diversity have
a venerable history, but those of phenotypic diversity (hereafter
\emph{disparity}) are a comparatively more recent phenomenon. Originally
defined as ``multidimensional morphological dissimilarity at a
macroevolutionary scale'' (Runnegar 1987, Gould
\href{https://paperpile.com/c/sTGYvp/Uns3}{2000}), this concept of
disparity emerged from attempts by palaeobiologists to characterise the
evolutionary origin of animal bodyplans and from attempts by comparative
developmental biologists to provide causal explanations for their
emergence. However, analyses of ``disparity'' have since expanded into
comparative biology as a means of capturing the effect of intrinsic and
extrinsic causal agents in morphological evolution. Typically, methods
to capture disparity are based on multidimensional spaces where each
dimension represents an aspect of morphological variation (a trait) and
biological observations (taxa) can be placed in this space based on
their trait values. Such multidimensional spaces (or morphospaces) can
then be used to tackle a diverse array of questions that can be grouped
into four main (non-mutually exclusive) classes (Fig. 1):

(1) \emph{Descriptive disparity.} The pioneering studies of disparity
characterised the shapes of organisms and how they differed among groups
\href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Foote 1995; Wills,
Briggs, and Fortey 1994)}. These studies consist of describing
multidimensional patterns in the diversity of morphological traits,
addressing questions such as: why are some morphological trait
combinations more common than others and what are the biological (or
mathematical) properties of the resulting morphospace?
\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Foote 1995; Raup
1961; Gerber 2017)}.

(2) \emph{Disparity-through-time.} This approach investigates how the
morphologies of organisms have changed on a temporal scale, focussing on
the disparity of taxa in particular time intervals or slices. This
approach has been used widely in palaeobiology to answer a range of
macroevolutionary questions, such as: how does disparity accumulate over
the history of a clade
\href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper
2018; Wright 2017)}, or how does disparity change leading up to and
across mass extinction events
\href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}?

(3) \emph{Disparity and taxonomic diversity.} Morphological disparity
provides another perspective on biodiversity; high morphological
disparity represents a high diversity of morphologies (i.e. shapes or
body plans) and is, presumably, associated with high levels of
ecological and functional diversity (though see below). This makes
disparity an informative complement to diversity measures based on
species richness alone. Indeed, most studies that have investigated
disparity and taxonomic diversity support an effective decoupling
\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey,
Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and
Neige 2007)}. This approach has been used to investigate whether some
groups are more successful than others in their exploration of new
evolutionary strategies
\href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.

(4) \emph{Disparity as a proxy for ecology.} The disparity of a group
can be used as a proxy for either the functional role it plays within an
ecosystem or its ecological niche. This approach assumes that groups
with high disparity are also likely to be functionally and ecologically
diverse, and that groups found in similar regions of shape space will
have similar functional and ecological roles
\href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010;
Pierce, Angielczyk, and Rayfield 2008; P. S. L. Anderson et al. 2011)}.
The links between form and function, however, are not always clear cut.
Traits can be linked to multiple functions and multiple functions can be
linked to a single trait
\href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al. 2005)}.
This approach has been used to investigate hypotheses of competitive
replacement (Stephen L. Brusatte et al. 2008) and changes in ecosystem
function during and after mass extinctions
\href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}. It is
particularly common in palaeobiology where it is not possible to
directly observe the ecological or functional characteristics of extinct
species \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.}
2005\href{https://paperpile.com/c/sTGYvp/Ejzr}{)}.

(5) Disparity and evo-devo. Morphological disparity blalbalbal

Fundamental insights into evolutionary biology have been elicited from
these four types of analysis. One of the most important insights is the
discovery that morphological disparity is often greatest early in the
evolutionary history of clades
\href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and
Wills 2013; Foote 1997;}
\href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that
the capacity for evolutionary innovation wanes with clade age, which
some have argued reflects the evolutionary assembly of gene regulatory
networks that constrain later fundamental change
\href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and
Wills 2013; Erwin 2011)}. However, this example highlights one of the
most challenging problems confronting researchers who are attempting,
increasingly, to obtain general insights from multiple independent
studies: can the insights gained from studies using the diversity of
methods, approaches and data types employed be considered equivalent?

In attempting to answer this question, we review current methods and
highlight their limitations, as part of a more general attempt to
propose best practice guidelines for studies of disparity . We first
discuss the appropriate data required for characterising disparity, then
review various challenging aspects of these approaches. Throughout, it
is important to remember these tools should always be used in the
context of a specific scientific question, as this will drive data and
methodological choices at every stage of the process.

\includegraphics[width=6.5in,height=4.94444in]{media/image1.png}

\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main
types of disparity analysis. \textbf{Descriptive disparity} focuses on
describing the features of morphospace occupation;
\textbf{disparity-through-time} investigates the evolution of the
morphospace through time including the effect of extinction events;
\textbf{disparity and taxonomic diversity} compares different measures
of biodiversity ; \textbf{disparity as a proxy for ecology} uses
disparity as a proxy for the ecological or functional role of a group.
These categories are not independent and many disparity studies will
cover more than one.}

\hypertarget{data-and-disparity}{%
\section{2. Data and disparity }\label{data-and-disparity}}

Disparity analyses are based on traits, but traits can be characterised
in a number of ways: (1) discrete ``cladistic'' characters, e.g. coding
the absence or presence of features or a discrete characteristic of a
trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
(2) continuous measurements of features (e.g., lengths)
\href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al.
2011)}; or (3) more mathematical descriptors from geometric morphometric
landmark data (e.g. Procrustes coordinates)
\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, and
Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote
1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al. 2018)})
(Fig. 2). None of these approaches are superior, but they may be more or
less well-suited to characterising traits under comparison and to the
question being asked of those traits (Hetherington et al 2015; Hopkins
2017).

For example, if investigating variation of bat wing shapes, both
homologous landmarks and continuous measurements of bones may be
appropriate to capture patterns of wing variation. However, if the
question focuses on comparing wings between bats and birds, different
measurements might be more appropriate depending on the specific
question: for example if the focus is on wing function, i.e. whether the
aerodynamic properties of wings vary within bats or between bats and
birds, the traits collected should reflect these aerodynamic properties
(e.g. wingspan, aspect ratio, etc.). However, if the focus is on
convergence between different bats and birds, it would be preferable to
use traits that have facilitated flight in both groups (e.g. digit
length, integumentary system, etc.). Where there is any doubt about the
appropriate traits to choose, it may be preferable to use different
kinds of data for the same feature to determine whether they capture the
same pattern of disparity.

The points above assume that researchers are collecting their own data
for disparity analyses, but often this is not the case. Discrete
character data are commonly recycled from phylogenetic studies
(e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline
et al. 2018}). This is an efficient approach to character sampling, but
it may artifactually increase disparity between phylogenetically
distinct groups because phylogenetic characters are often collected to
discriminate among groups
\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}. This needs to
be considered when interpreting results, especially as synapomorphies
will naturally lead to an apparent shift or increase in disparity when
new clades appear. Furthermore, many datasets are limited to subsets of
anatomy that are at least implicit samples of overall anatomy, but
explicit tests of this assumption have shown that different aspects of
morphology can exhibit different patterns of disparity
\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}.
This effect of anatomical part on disparity patterns can be especially
challenging when working on datasets where the available data has
non-random missing anatomical parts, such as the absence of soft tissue
in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline
et al. 2018)}.

Trait data suffer from the same shortcomings as most biological datasets
-- data can be missing, non-overlapping, hierarchical, inapplicable,
ambiguous, polymorphic, correlated, or there may be an insufficient
sample size \href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau,
Guillerme, and Smith 2017};
\href{https://paperpile.com/c/sTGYvp/yO2t}{Palci and Lee 2018)}.
Biological phenomena such as allometry and sexual dimorphism may also
influence trait data. More mundanely, data collection is constrained by
the time and money available, making collating a ``perfect'' dataset
difficult. Ultimately, disparity analyses characterise the data, and
subsamples of the universe of possible data may not have the power to
uncover holistic patterns of disparity. Therefore, trait data should be
collected with the question in mind, or the question asked should be
tailored to the limits of the data available.

\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.

\emph{Figure data {[}figure\_data.pdf{]}: Major routes to obtain
morphological data for disparity analyses. Data can be collected as
discrete trait observations (e.g. presence or absence data) or as
continuous data. Continuous data can be collected by various methods
including linear measurements and landmark coordinates or contours
(curves). These measurements can then be mathematically transformed
(logarithm transforms, scaling, Procrustes superimposition, elliptic
Fourier transforms, etc.). Regardless of the method, data collection
produces a trait matrix where the observed traits constitute columns and
the studied elements (generally taxa or OTUs) the rows.}

\hypertarget{disparity-methods}{%
\section{Disparity methods}\label{disparity-methods}}

Once suitable trait data have been collected, the design of the
disparity analysis itself needs to be considered. Study design
encompasses several key aspects including (a) the difficulty of dealing
with multidimensional data; (b) the variety of ordination (dimension
reduction) techniques available and their limitations; (c) the metrics
used to summarise the relative disparity of groups; (d) the methods used
for hypothesis testing within the disparity analysis framework; and (e)
ancestral state estimations in disparity-through-time analyses. We
consider these aspects in turn below.

\hypertarget{a-multidimensionality}{%
\subsection{(a) Multidimensionality}\label{a-multidimensionality}}

Disparity analyses often use ordination techniques for dimension reduction.
These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g. principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}. %TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al. 2018;
Gerber 2017)}.
Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e. contrary to a PCA, the first few axis do not encompass most variance of the dataset).
Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e. the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.

\hypertarget{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}{%
\subsection{(b) To ordinate or not to ordinate - that is the
(multidimensional)
question}\label{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}}

Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
